---
title: "The Effect of Climate on Aggregate Crop Choice: Bayesian Modeling with R and STAN"
output: html_notebook
bibliography: "`r here::here('outputs/library.bib')`"
---

# Overview of Berry-Levinsohn-Paker (BLP) / aggregate random coefficients logit approach

The problem: __we don't have individual-level data and choices aren't discrete.__

- If discrete choices: multinomial logit model
- We just have aggregate crop "shares" as a share of total county acres, so use a __BLP / aggregate random coefficients logit__ model.

## Model

Following @Arriagada2005, production of crop $i$ in county $g$ is undertaken with some production function 

$$q_{ig} = f_i(x_{ig}, z_g),$$ 

where $x_{ig}$ are variable inputs and $z_g$ are exogenous inputs that include weather.

Given crop price $p_i$, the farmer chooses $x_{ig}$ to maximize profits such that

$$\Pi_{ig}(p_i, z_g) = \max \limits_{x_{ig}} \{ p_i q_i (p_i, z_g) - c_i(p_i, z_g) \bar{c}_i\},$$

where $\bar{c}_i$ is the mean cost expenditure for crop $i$.

Given exogenous shocks $z_g$, mean costs $\bar{c}_i$, and prices $p_i$, the farmer chooses the crop that yields the highest profit in that county

$$\Pi_{*g}(p, \bar{c}, z_g) = \max \limits_i \{ \Pi_{ig}(p_i, \bar{c}_i, z_g) \}.$$

Unobserved attributes known to the farmer for crop $i$ in county $g$ and year $t$ are represented by $\alpha_{igt} = \alpha_i + \Delta \alpha_{igt}.$ Then profits are given by

$$\Pi_{igt}(\cdot) = \alpha_i + \Delta \alpha_{igt} + \beta(z_{gt}, \phi_{g}) p_{it} + \delta (z_{gt}, \phi_{g}) \bar{c}_{it} + \varepsilon_{igt},$$

where $\varepsilon_{igt}$ is a mean-zero stochastic term and $\beta(\cdot)$ and $\delta(\cdot)$ are productivity parameters for land in county $g$ given observed and unobserved characteristics $z_{gt}$ and $\phi_g$.

$\beta$ and $\delta$ are random coefficients given by

$$(\beta, \delta) = B + \Lambda z_{gt} + \Sigma \phi_{g}, \phi_g \sim N(0, I)$$

## Need an outside option

An outside option serves as a reference point against which the profit from all crops is measured. Here it is natural to let this be farmland that is fallow/unplanted.


# Bayesian modeling

## Workflow

1. Define the model
2. Create fake data with known parameters
3. Create a model to estimate those parameters
4. Check that it correctly estimates
5. Apply model to estimate parameters for real data

## Useful links

- [STAN Documentation and tutorials]()


## Fake data params

Let's set some standard parameters for the fake data we'll generate:

```{r}
library(here)   # always know the root of your project
library(data.table)
library(gtools) # for the Dirichlet distribution
library(rstan)  # running STAN models
library(beepr)  # plays noise when you're done
options(mc.cores = parallel::detectCores())

set.seed(57) #reproducable randomness

# Dimensions of the data. 
n_ind <- 500     # 500 fake farmers in each county
n_crops <- 3     # 5 crops
n_counties <- 50 # 10 counties
k <- 2           # covariates
```

## A simple multinomial model

Let's start with just a simple multinomial choice model. Imagine that each county chooses it's crop mix according to the same set of probabilities.

```{r}
### Generate Fake Data ---------------------------------------------------------
# total acres for each county
acres <- rpois(n_counties, 213000) #value comes from the mean acres of all counties

# probability of selecting each crop
theta <- rdirichlet(1, rnorm(n_crops+1, 5, 1))  #use lognormal because can't have negative probabilities of selection.

# outcomes
crop_acres <- t(sapply(1:n_counties, function(g) {
  rmultinom(1, acres[g], theta) 
  }))


### STAN MODEL -----------------------------------------------------------------
simplest_stan <- 
'
// Data block, determines data inputs.
data {
  int n_counties;
  int n_crops;
  int acres[n_counties, n_crops];
}

// Parameter block, determines what parameters are allowed
parameters {
  simplex[n_crops] theta; //probability of each crop being selected
}

// Model block, specifies model and priors
model {
  // prior on theta
  theta ~ beta(2,1); 
  
  // model
  for(g in 1:n_counties) {
    acres[g] ~ multinomial(theta);
  }
}
'

### Build the model and estimate parameter values
fake_data_list <- list(n_counties = n_counties, n_crops = n_crops+1, acres = crop_acres)

model1 <- stan_model(model_code = simplest_stan); beepr::beep("fanfare")

# use optimization instead of sampling for model development
res_optim <- optimizing(model1, data = fake_data_list)

# use for actual production run
res_sample <- stan(model_code = simplest_stan, data = fake_data_list, chains = 3, iter = 1000, warmup = 500, thin = 10)

# Look at chain convergence, pretty good!
traceplot(res_sample, pars = c("theta"), inc_warmup = TRUE)
```


## Okay, now let's make things a little more complex...

The probability of selecting each crop likely depends on some county characteristics such as weather, infrastructure, etc... Let's represent those characteristics by $X$. Also let $theta$ be a function of $X$, so that $\theta \sim \mbox{Dirichlet}(X \beta')$.

```{r}
### Generate More Fake Data ---------------------------------------------------------
X <- matrix(rlnorm(n_counties*k, 0, 1), n_counties, k) # county covariates
X_long <- matrix(sapply(1:n_counties, function(i) { 
  matrix(rep(X[i,], each = n_crops), nrow = n_crops) 
  }), ncol = k)
  

# probability of selecting each crop
beta <- rnorm(k, 0, 1) #coefficients that determine theta
alpha <- matrix(X_long %*% beta+10, n_counties, n_crops, byrow = TRUE)
theta <- t(sapply(1:n_counties, function(i) { rdirichlet(1, alpha[i,]) }))

# outcomes
crop_acres <- t(sapply(1:n_counties, function(g) {
  rmultinom(1, acres[g], theta[g,]) 
  }))


### STAN MODEL -----------------------------------------------------------------
fake_data_list <- list(n_counties = n_counties, n_crops = n_crops, k = k, X = X_long, acres = crop_acres)

model2 <- stan_model(file = here("src/stan/dirichlet_multinomial.stan")); beepr::beep("fanfare")

# use optimization instead of sampling for model development
res_optim <- optimizing(model2, data = fake_data_list)

# use for actual production run
res_sample <- stan(file = here("src/stan/dirichlet_multinomial.stan"), data = fake_data_list, chains = 3, iter = 1000, warmup = 500, thin = 10)

# Look at chain convergence, pretty good!
traceplot(res_sample, pars = c("theta"), inc_warmup = TRUE)
```

## Recovering the coefficients that determine shares

We want to access our betas, and later we want to have more control over how they are determined

```{r}
# Generate theta shares directly
beta <- matrix(rnorm(n_crops*k, 0, 1), n_crops, k)
theta <- t(sapply(1:n_counties, function(i) { 
  alpha = exp(rowSums(X_long[((i-1)*n_crops + 1):(i*n_crops),] * beta))
  alpha/(1 + sum(alpha))
  }))
theta <- cbind(theta, 1 - rowSums(theta))

# outcomes
crop_acres <- t(sapply(1:n_counties, function(g) {
  rmultinom(1, acres[g], theta[g,]) 
  }))


### STAN MODEL -----------------------------------------------------------------
fake_data_list <- list(n_counties = n_counties, n_crops = n_crops, k = k, X = X_long, acres = crop_acres)

model3 <- stan_model(file = here("src/stan/multinomial2.stan")); beepr::beep("fanfare")

# use optimization instead of sampling for model development
res_optim <- optimizing(model3, data = fake_data_list)
colMeans(beta)

# use for actual production run
res_sample <- stan(file = here("src/stan/multinomial2.stan"), data = fake_data_list, chains = 3, iter = 1000, warmup = 500, thin = 10)

# Look at chain convergence, pretty good!
traceplot(res_sample, pars = c("beta"), inc_warmup = TRUE)
```

## Now for some real data!

```{r}
d <- readRDS(here("data/cropmix_sample.rds"))
d[, `:=`(
  n_gdd = gdd/max(gdd),
  n_rain = rain/max(rain),
  n_days = days/max(days),
  n_elev = elev/max(elev))]

# Just a subset of crops
#crop_names <- unique(d$crop)
crop_names <- c("Alfalfa", "Almonds", "Apples", "Asparagus", "Barley", "Corn", "Pecans", "Spring Wheat", "Winter Wheat")
d2 <- d[crop %in% crop_names]
n_counties <- length(unique(d2$fips))
n_crops <- length(crop_names)

# X matrix
x_vars <- c("n_gdd", "n_rain", "n_days", "clay", "silt", "sand", "n_elev")
k <- length(x_vars) + 1
X <- cbind(1, as.matrix(d2[, x_vars, with=F]))

# Acres
fallow <- d[crop == "Fallow/Idle Cropland"]$crop_acres
crop_acres <- matrix(round(c(fallow, d2$crop_acres)), n_counties, n_crops+1) 

# fit using a non-bayesian multinomial model
X1 <- unique(X) #no duplicate rows
m <- nnet::multinom(crop_acres ~ X1 - 1)

# Data for our model
data_list <- list(n_counties = n_counties, n_crops = n_crops, k = k, X = X, acres = crop_acres)

# model
model <- stan_model(file = here("src/stan/multinomial3.stan")); beepr::beep("fanfare")
expose_stan_functions(here("src/stan/multinomial3.stan"))

# use optimization instead of sampling for model development
res_optim <- optimizing(model, data = data_list)

# use for actual production run
res_sample <- stan(file = here("src/stan/multinomial3.stan"), data = data_list, chains = 3, iter = 1000, warmup = 500, thin = 10)

# Look at chain convergence
traceplot(res_sample, pars = c("beta"), inc_warmup = TRUE)

# shinystan
shinystan::launch_shinystan(res_sample)
```

## References

