---
title: "The Effect of Climate on Aggregate Crop Choice: Bayesian Modeling with R and STAN"
output: html_notebook
bibliography: "`r here::here('outputs/library.bib')`"
---

We've been trying to get at this question

> how is projected climate in 2040 likely to affect economic productivity of agriculture and how are producers likely to adapt to maximize their profits?

- One method is to do a cross-sectional analysis of counties as in @MendelsohnNordhausShaw1994.
- Another is a panel/fixed-effects model such as @DeschenesGreenstone2007.
- A third is a "long differences" approach as in @BurkeEmerick2016.
- Our approach is to use matching @PotterETAL2018. Here are [slides](https://potterzot.com/research/agClimateImpact/agClimateImpact_SES2018.html).

In that method we take each county in 2040 and find the 2012 county climate that most matches it. The difference between the outcome variable in these two counties is the estimate of the "effect" of the projected climate in 2040. 

We can do that with crop mixes also...

# But can we also look at crop switching directly through a multinomial choice model?

There are 109 different crops crown so probably not, and we don't have individual producer-level information on crop choice.

__But__, we do have aggregate county-level crop mix.

Enter the @BerryETAL1995 (BLP) / aggregate random coefficients logit model


## Model

Following @Arriagada2005, production of crop $i$ in county $g$ is undertaken with some production function 

$$q_{ig} = f_i(x_{ig}, z_g),$$ 

where $x_{ig}$ are variable inputs and $z_g$ are exogenous inputs that include weather.

Given crop price $p_i$, the farmer chooses $x_{ig}$ to maximize profits such that

$$\Pi_{ig}(p_i, z_g) = \max \limits_{x_{ig}} \{ p_i q_i (p_i, z_g) - c_i(p_i, z_g) \bar{c}_i\},$$

where $\bar{c}_i$ is the mean cost expenditure for crop $i$.

Given exogenous shocks $z_g$, mean costs $\bar{c}_i$, and prices $p_i$, the farmer chooses the crop that yields the highest profit in that county

$$\Pi_{*g}(p, \bar{c}, z_g) = \max \limits_i \{ \Pi_{ig}(p_i, \bar{c}_i, z_g) \}.$$

Unobserved attributes known to the farmer for crop $i$ in county $g$ and year $t$ are represented by $\alpha_{igt} = \alpha_i + \Delta \alpha_{igt}.$ Then profits are given by

$$\Pi_{igt}(\cdot) = \alpha_i + \Delta \alpha_{igt} + \beta(z_{gt}, \phi_{g}) p_{it} + \delta (z_{gt}, \phi_{g}) \bar{c}_{it} + \varepsilon_{igt},$$

where $\varepsilon_{igt}$ is a mean-zero stochastic term and $\beta(\cdot)$ and $\delta(\cdot)$ are productivity parameters for land in county $g$ given observed and unobserved characteristics $z_{gt}$ and $\phi_g$.

$\beta$ and $\delta$ are random coefficients given by

$$(\beta, \delta) = B + \Lambda z_{gt} + \Sigma \phi_{g}, \phi_g \sim N(0, I)$$

## Need an outside option

An outside option serves as a reference point against which the profit from all crops is measured. Here it is natural to let this be farmland that is fallow/unplanted.


# Bayesian modeling

## Workflow

1. Define the model
2. Create fake data with known parameters
3. Create a model to estimate those parameters
4. Check that it correctly estimates
5. Apply model to estimate parameters for real data

## Useful links

- [STAN Documentation and tutorials]()


## Fake data params

Let's set some standard parameters for the fake data we'll generate:

```{r}
library(here)   # always know the root of your project
library(data.table)
library(gtools) # for the Dirichlet distribution
library(rstan)  # running STAN models
library(shinystan) # for model checking
library(beepr)  # plays noise when you're done
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)

set.seed(57) #reproducable randomness

# Dimensions of the data. 
n_ind <- 500     # 500 fake farmers in each county
n_crops <- 3     # 5 crops
n_counties <- 50 # 10 counties
k <- 2           # covariates
```

## A simple multinomial model

Let's start with just a simple multinomial choice model. Imagine that each county chooses it's crop mix according to the same set of probabilities.

```{r}
### Generate Fake Data ---------------------------------------------------------
# total acres for each county
acres <- rpois(n_counties, 213000) #value comes from the mean acres of all counties

# probability of selecting each crop
theta <- rdirichlet(1, rnorm(n_crops+1, 5, 1))  #use lognormal because can't have negative probabilities of selection.

# outcomes
crop_acres <- t(sapply(1:n_counties, function(g) {
  rmultinom(1, acres[g], theta) 
  }))


### STAN MODEL -----------------------------------------------------------------
simplest_stan <- 
'
// Data block, determines data inputs.
data {
  int n_counties;
  int n_crops;
  int acres[n_counties, n_crops];
}

// Parameter block, determines what parameters are allowed
parameters {
  simplex[n_crops] theta; //probability of each crop being selected
}

// Model block, specifies model and priors
model {
  // prior on theta
  theta ~ beta(2,1); 
  
  // model
  for(g in 1:n_counties) {
    acres[g] ~ multinomial(theta);
  }
}
'

### Build the model and estimate parameter values
fake_data_list <- list(n_counties = n_counties, n_crops = n_crops+1, acres = crop_acres)

model1 <- stan_model(model_code = simplest_stan); beepr::beep("fanfare")

# use optimization instead of sampling for model development
res_optim <- optimizing(model1, data = fake_data_list)

# use for actual production run
res_sample <- stan(model_code = simplest_stan, data = fake_data_list, chains = 3, iter = 1000, warmup = 500, thin = 10)

# Look at chain convergence, pretty good!
traceplot(res_sample, pars = c("theta"), inc_warmup = TRUE)

shinystan::launch_shinystan(res_sample)
```


## Okay, now let's make things a little more complex...

The probability of selecting each crop likely depends on some county characteristics such as weather, infrastructure, etc... Let's represent those characteristics by $X$. Also let $theta$ be a function of $X$, so that $\theta \sim \mbox{Dirichlet}(X \beta')$.

```{r}
### Generate More Fake Data ---------------------------------------------------------
X <- matrix(rlnorm(n_counties*k, 0, 1), n_counties, k) # county covariates
X_long <- matrix(sapply(1:n_counties, function(i) { 
  matrix(rep(X[i,], each = n_crops), nrow = n_crops) 
  }), ncol = k)
  

# probability of selecting each crop
beta <- rnorm(k, 0, 1) #coefficients that determine theta
alpha <- matrix(X_long %*% beta+15, n_counties, n_crops, byrow = TRUE)
theta <- t(sapply(1:n_counties, function(i) { rdirichlet(1, alpha[i,]) }))

# outcomes
crop_acres <- t(sapply(1:n_counties, function(g) {
  rmultinom(1, acres[g], theta[g,]) 
  }))


### STAN MODEL -----------------------------------------------------------------
fake_data_list <- list(n_counties = n_counties, n_crops = n_crops, k = k, X = X_long, acres = crop_acres)

model2 <- stan_model(file = here("src/stan/dirichlet_multinomial.stan")); beepr::beep("fanfare")

# use optimization instead of sampling for model development
res_optim <- optimizing(model2, data = fake_data_list)

# use for actual production run
# ~20 seconds
res_sample <- stan(file = here("src/stan/dirichlet_multinomial.stan"), data = fake_data_list, chains = 3, iter = 1000, warmup = 500, thin = 10)
pairs(res_sample)

# with adapt_delta
res_sample <- stan(file = here("src/stan/dirichlet_multinomial.stan"), data = fake_data_list, chains = 3, iter = 1000, warmup = 500, thin = 10, control = list(adapt_delta = 0.99))

# Look at chain convergence, pretty good!
traceplot(res_sample, pars = c("theta"), inc_warmup = TRUE)
```

## Recovering the coefficients that determine shares

We want to access our betas, and later we want to have more control over how they are determined

```{r}
# Generate theta shares directly
beta <- matrix(rnorm(n_crops*k, 0, 1), n_crops, k)
theta <- t(sapply(1:n_counties, function(i) { 
  alpha = exp(rowSums(X_long[((i-1)*n_crops + 1):(i*n_crops),] * beta))
  alpha/(1 + sum(alpha))
  }))
theta <- cbind(theta, 1 - rowSums(theta))

# outcomes
crop_acres <- t(sapply(1:n_counties, function(g) {
  rmultinom(1, acres[g], theta[g,]) 
  }))


### STAN MODEL -----------------------------------------------------------------
fake_data_list <- list(n_counties = n_counties, n_crops = n_crops, k = k, X = X_long, acres = crop_acres)

model3 <- stan_model(file = here("src/stan/multinomial2.stan")); beepr::beep("fanfare")

# use optimization instead of sampling for model development
res_optim <- optimizing(model3, data = fake_data_list)
colMeans(beta)

# use for actual production run
res_sample <- stan(file = here("src/stan/multinomial2.stan"), data = fake_data_list, chains = 3, iter = 1000, warmup = 500, thin = 10)

# Look at chain convergence, pretty good!
traceplot(res_sample, pars = c("beta"), inc_warmup = TRUE)
```

## Now for some real data!

```{r}
d <- readRDS(here("data/cropmix_sample.rds"))
d[, `:=`(
  n_gdd = gdd/max(gdd),
  n_rain = rain/max(rain),
  n_days = days/max(days),
  n_elev = elev/max(elev))]

# Just a subset of crops
#crop_names <- unique(d$crop)
crop_names <- c("Alfalfa", "Almonds", "Apples", "Asparagus", "Barley", "Corn", "Pecans", "Spring Wheat", "Winter Wheat")
d2 <- d[crop %in% crop_names]
n_counties <- length(unique(d2$fips))
n_crops <- length(crop_names)

# X matrix
x_vars <- c("n_gdd", "n_rain", "n_days", "clay", "silt", "sand", "n_elev")
k <- length(x_vars) + 1
X <- cbind(1, as.matrix(d2[, x_vars, with=F]))

# Acres
fallow <- d[crop == "Fallow/Idle Cropland"]$crop_acres
crop_acres <- matrix(round(c(fallow, d2$crop_acres)), n_counties, n_crops+1) 

# fit using a non-bayesian multinomial model
X1 <- unique(X) #no duplicate rows
m <- nnet::multinom(crop_acres ~ X1 - 1)

# Data for our model
data_list <- list(n_counties = n_counties, n_crops = n_crops, k = k, X = X, acres = crop_acres)

# model
model <- stan_model(file = here("src/stan/multinomial3.stan")); beepr::beep("fanfare")
expose_stan_functions(here("src/stan/multinomial3.stan"))

# use optimization instead of sampling for model development
res_optim <- optimizing(model, data = data_list)

# use for actual production run
res_sample <- stan(file = here("src/stan/multinomial3.stan"), data = data_list, 
                   chains = 3, iter = 1000, warmup = 500, thin = 10,
                   control = list(max_treedepth = 15))

# shinystan
shinystan::launch_shinystan(res_sample)
```

## References

